``Interestingness'' in Mathematical Theory Exploration

### 2019-06-02 PhDThesis/litreview_testing.tex
Headingy things
Thesis updates

\subsection{Property Checking}

Some prominent tools for mathematical theory exploration, like \qspec{}, were
originally developed in the context of \emph{property checking} (although for
\emph{discovering} software properties, rather than checking them), and only
later applied more directly to fields like formal mathematics (e.g. via the
\hipster{} tool for Isabelle/HOL). In particular, the logical domain used by
\qspec{} (inductive logic with higher-order functions) comes directly from its
application to Haskell; and its search algorithm makes heavy use of the
\qcheck{} property checker. The boundary between these two fields is hence
rather fuzzy, and cross-pollination of techniques is a compelling target for
research.

@@

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \hs{forAll}), or by
choosing a default generator for types implementing its \hs{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
\cite{TODO} \fi).

@@

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \hs{shrink} function along with their generator,
which is given any counterexample found during testing and returns a list of
smaller values. In the above example we might shrink a \hs{Node} by returning
each sub-tree, and return an empty list when asked to shrink a \hs{Leaf}. If one
of these smaller values is also a counterexample, it too is shrunk and the
process repeats, otherwise it is discarded and the next element of the list is
tried. When a value cannot be shrunk any more, or if none of its shrunken
versions are counterexamples, that value is presented to the user in the hope
that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumeration function or using its
\hs{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
\hs{Tree} example this might check \hs{Leaf}, then \hs{Node Leaf
  Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
order of ``size'' ensures that counterexamples are always small, that tests are
not repeated with the same values and that memory is not exhausted (since the
stopping condition will trigger long before the enumeration reaches such large
values). The disadvantage is that large inputs (which exercise properties,
especially recursive ones, the most) may occur so late in the enumeration that
testing never reaches them.

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \hs{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which refines the
previous value and tries again. For example, a lazily generated \hs{Tree} may
initially be equivalent to $\bot$ (an undefined value). If the property
aborts, it will be run again with each possible refinement of $\bot$: in this
case inputs equivalent to \hs{Leaf} and \hs{Node}~$\bot_1$~$\bot_2$. If
the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
give inputs \hs{Node Leaf}~$\bot_2$ and
\hs{Node (Node}~$\bot_3$~$\bot_4$\hs{)}~$\bot_2$ (and likewise if $\bot_2$
had caused the exception). This refinement continues until the property no
longer aborts or until a predefined depth limit is reached. If the property
returns a true value when given such partially-defined input then it must
succeed for \emph{all} refinements of that input (since the output was
determined without evaluating the undefined sub-terms). If the property returns
a false value, the input is a \emph{minimal} counterexample, since it only
specifies those sub-terms which are required to violate the property; any
undefined sub-terms are irrelevant to the failure.

\textsc{LazySmallCheck}'s automatic elaboration of free variables is similar in
operation to the use of search procedures in logic programming languages like
Kanren, Prolog (which has its own \textsc{QuickCheck} port) and Curry (a
functional logic language, whose \textsc{EasyCheck} framework explicitly
unifies the concepts of generators and logic
variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
is also reminiscent of \emph{symbolic execution}~\cite{king1976symbolic}, a form
of \emph{abstract interpretation} where unknown runtime values are represented
by opaque symbols and computation proceeds to arrive at a result, which may be a
complicated expression involving many symbols.

>>>>

\section{Property Checking}

Some prominent tools for mathematical theory exploration, like \quickspec{},
were originally developed in the context of \emph{property checking} (although
for \emph{discovering} software properties, rather than checking them), and only
later applied more directly to fields like formal mathematics (e.g. via the
\hipster{} tool for Isabelle/HOL). In particular, the logical domain used by
\quickspec{} (inductive logic with higher-order functions) comes directly from
its application to Haskell; and its search algorithm makes heavy use of the
\quickcheck{} property checker. The boundary between these two fields is hence
rather fuzzy, and cross-pollination of techniques is a compelling target for
research.

@@

\quickcheck{}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \quickcheck{} generates input data \emph{randomly},
either using explicit generators (via a function called \hs{forAll}), or by
choosing a default generator for types implementing its \hs{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
\cite{TODO} \fi).

@@

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\quickcheck{} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \hs{shrink} function along with their generator,
which is given any counterexample found during testing and returns a list of
smaller values. In the above example we might shrink a \hs{Node} by returning
each sub-tree, and return an empty list when asked to shrink a \hs{Leaf}. If one
of these smaller values is also a counterexample, it too is shrunk and the
process repeats, otherwise it is discarded and the next element of the list is
tried. When a value cannot be shrunk any more, or if none of its shrunken
versions are counterexamples, that value is presented to the user in the hope
that it is simple enough to understand.

The \smallcheck{} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumeration function or using its
\hs{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
\hs{Tree} example this might check \hs{Leaf}, then \hs{Node Leaf
  Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
order of ``size'' ensures that counterexamples are always small, that tests are
not repeated with the same values and that memory is not exhausted (since the
stopping condition will trigger long before the enumeration reaches such large
values). The disadvantage is that large inputs (which exercise properties,
especially recursive ones, the most) may occur so late in the enumeration that
testing never reaches them.

The \lazysmallcheck{} framework is similar in spirit to
\smallcheck{}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \hs{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \lazysmallcheck{} which refines the
previous value and tries again. For example, a lazily generated \hs{Tree} may
initially be equivalent to $\bot$ (an undefined value). If the property
aborts, it will be run again with each possible refinement of $\bot$: in this
case inputs equivalent to \hs{Leaf} and \hs{Node}~$\bot_1$~$\bot_2$. If
the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
give inputs \hs{Node Leaf}~$\bot_2$ and
\hs{Node (Node}~$\bot_3$~$\bot_4$\hs{)}~$\bot_2$ (and likewise if $\bot_2$
had caused the exception). This refinement continues until the property no
longer aborts or until a predefined depth limit is reached. If the property
returns a true value when given such partially-defined input then it must
succeed for \emph{all} refinements of that input (since the output was
determined without evaluating the undefined sub-terms). If the property returns
a false value, the input is a \emph{minimal} counterexample, since it only
specifies those sub-terms which are required to violate the property; any
undefined sub-terms are irrelevant to the failure.

\lazysmallcheck{}'s automatic elaboration of free variables is similar in
operation to the use of search procedures in logic programming languages like
Kanren, Prolog (which has its own \quickcheck{} port) and Curry (a
functional logic language, whose \textsc{EasyCheck} framework explicitly
unifies the concepts of generators and logic
variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
is also reminiscent of \emph{symbolic execution}~\cite{king1976symbolic}, a form
of \emph{abstract interpretation} where unknown runtime values are represented
by opaque symbols and computation proceeds to arrive at a result, which may be a
complicated expression involving many symbols.

### 2019-06-02 PhDThesis/ml4hs.tex
Thesis updates

We provide an implementation of our recurrent clustering algorithm in a tool
called \textsc{ML4HS}, which consists of a loose collection of components shown
in Figure \ref{fig:ml4hs}. This arrangement makes it easy to swap out parts for
experimentation. In the following, we describe the custom components in the
order they appear in the diagram.

@@

  \begin{tikzpicture}[node distance=2cm]
    \node [block] (hackage) {Hackage};

@@

    \node [block, below of=clustering, fill=shade] (mlspec) {
      \begin{tikzpicture}[node distance = 2cm, auto]
        \node (mlspeclabel) {MLSpec};
        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {QuickSpec};
      \end{tikzpicture}
    };

@@

Since we are interested in comparing definitions based on the terms they
reference, building our framework on top of GHC seems like a promising
approach. Indeed, \hspec{} already invokes GHC's API to obtain the definitions
of Haskell functions, in order to transform them into a form suitable for ATP
systems. However, our initial experiments showed that this technique is too
fragile for use on many real Haskell projects.

@@

\subsection{\textsc{MLSpec}}
\label{sec:mlspec}

We cannot supply these clusters as-is to \qspec{}, since it must be provided
with a \emph{signature}. These are constructed by our \textsc{MLSpec} tool,
using information from the \textsc{AstPlugin} database. Tasks performed by
\textsc{MLSpec} include:

\begin{itemize}
\item{Monomorphising}: Given values of polymorphic type, e.g. \hs{safeHead ::
    forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based
  system like \qspec{} is unable to evaluate these expressions without
  instantiating the variable \hs{t} to a specific type. Such an instantiation is
  called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on
  previous work in \qcheck{} by attempting to instantiate all type variables to
  \hs{Integer}. We discard those cases where this is invalid, such as variable
  \emph{type constructors} (e.g. \hs{forall c. c Bool -> c Bool}) or
  incompatible class constraints (e.g. \hs{forall t. IsString t => t}).

@@

\item{Variable definition}: Once a \qspec{} theory has been defined containing
  all of the given terms, we inspect the types it references and append three
  variables for each to the theory (enough to discover laws such as
  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
  search).

\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
  that all required packages and modules must be provided up-front, usually by
  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
  \emph{dynamically}, depending on the cluster information it is given, we do
  not know what packages it may need. To work around this problem,
  \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have
  built called \texttt{nix-eval}. This provides an \texttt{eval} function, like
  those commonly found in dynamic languages such as Python and Javascript, for
  evaluating Haskell expressions. The key feature of \texttt{nix-eval} is that
  these Haskell expressions may reference packages that are not installed on the
  system. When such expressions are evaluated, these packages will be
  automatically downloaded and installed into a sandbox using the Nix package
  manager, and GHC will be invoked in this sandbox to perform the evaluation.
\end{itemize}

>>>>

We provide an implementation of our recurrent clustering algorithm in a tool
called \mlforhs{}, which consists of a loose collection of components shown
in Figure \ref{fig:ml4hs}. This arrangement makes it easy to swap out parts for
experimentation. In the following, we describe the custom components in the
order they appear in the diagram.

@@

  \begin{tikzpicture}[node distance=2cm]
    \node [block] (hackage) {\hackage{}};

@@

    \node [block, below of=clustering, fill=shade] (mlspec) {
      \begin{tikzpicture}[node distance = 2cm, auto]
        \node (mlspeclabel) {\mlspec{}};
        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {\quickspec{}};
      \end{tikzpicture}
    };

@@

Since we are interested in comparing definitions based on the terms they
reference, building our framework on top of GHC seems like a promising
approach. Indeed, \hipspec{} already invokes GHC's API to obtain the definitions
of Haskell functions, in order to transform them into a form suitable for ATP
systems. However, our initial experiments showed that this technique is too
fragile for use on many real Haskell projects.

@@

\subsection{\mlspec{}}
\label{sec:mlspec}

We cannot supply these clusters as-is to \quickspec{}, since it must be provided
with a \emph{signature}. These are constructed by our \mlspec{} tool,
using information from the \astplugin{} database. Tasks performed by
\mlspec{} include:

\begin{itemize}
\item{Monomorphising}: Given values of polymorphic type, e.g. \hs{safeHead ::
    forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based
  system like \quickspec{} is unable to evaluate these expressions without
  instantiating the variable \hs{t} to a specific type. Such an instantiation is
  called \emph{monomorphising}, and in the case of \mlspec{} we build on
  previous work in \quickcheck{} by attempting to instantiate all type variables to
  \hs{Integer}. We discard those cases where this is invalid, such as variable
  \emph{type constructors} (e.g. \hs{forall c. c Bool -> c Bool}) or
  incompatible class constraints (e.g. \hs{forall t. IsString t => t}).

@@

\item{Variable definition}: Once a \quickspec{} theory has been defined containing
  all of the given terms, we inspect the types it references and append three
  variables for each to the theory (enough to discover laws such as
  associativity, but not too many to overflow the limit of \quickspec{}'s exhaustive
  search).

\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
  that all required packages and modules must be provided up-front, usually by
  specification in a Cabal file. Since \mlspec{} builds signatures
  \emph{dynamically}, depending on the cluster information it is given, we do
  not know what packages it may need. To work around this problem,
  \mlspec{} invokes \quickspec{} for each cluster using a library we have
  built called \texttt{nix-eval}. This provides an \texttt{eval} function, like
  those commonly found in dynamic languages such as Python and Javascript, for
  evaluating Haskell expressions. The key feature of \texttt{nix-eval} is that
  these Haskell expressions may reference packages that are not installed on the
  system. When such expressions are evaluated, these packages will be
  automatically downloaded and installed into a sandbox using the Nix package
  manager, and GHC will be invoked in this sandbox to perform the evaluation.
\end{itemize}

### 2019-06-02 PhDThesis/related_tests.tex
Headingy things
Thesis updates

Since we are working in the domain of Haskell programs, an abundant source of
statements are available in the form of \emph{tests}. For manually-written
tests, the effort required to write them implies that they must be of some
interest to their author. Many forms of test are \emph{not} suitable for our
purposes, such as \emph{unit tests} which are always trivially provable by
$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
the external environment. One reason to study Haskell is its widespread use of
\emph{property checking}, which does give us useful data. Many Haskell property
checkers exist, based on random testing (\textsc{QuickCheck}
\cite{claessen2011quickcheck} and \textsc{SmartCheck}
\cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck}
\cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck}
\cite{reich2013advances}) and logic programming (\textsc{SparseCheck}
\cite{sparsecheck}).

@@

Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of
2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
\texttt{blaze-html}.

>>>>

% TODO: This is not related work!

Since we are working in the domain of Haskell programs, an abundant source of
statements are available in the form of \emph{tests}. For manually-written
tests, the effort required to write them implies that they must be of some
interest to their author. Many forms of test are \emph{not} suitable for our
purposes, such as \emph{unit tests} which are always trivially provable by
$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
the external environment. One reason to study Haskell is its widespread use of
\emph{property checking}, which does give us useful data. Many Haskell property
checkers exist, based on random testing
(\quickcheck{}~\cite{claessen2011quickcheck} and
\smartcheck{}~\cite{pike2014smartcheck}), enumeration
(\smallcheck{}~\cite{runciman2008smallcheck}), observation
(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
(\sparsecheck{}~\cite{sparsecheck}).

@@

Here we consider the 10 most-downloaded packages from \hackage{} (as of
2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
\texttt{blaze-html}.
